Supplementary information - The Functional Therapeutic Chemical Classification System

### 2013-08-19 document.tex
final editing on the way
Signed-off-by: loopasam <samuel.croset@gmail.com>

\section{Source code}
The code behind the creation of the resource is entirely open and available 
at {{https://github.com/loopasam/ftc}}. The web application built on the top of the 
FTC can be find at {{https://www.ebi.ac.uk/chembl/ftc}} and the documentation can be 
accessed at {{https://github.com/loopasam/ftc/wiki}}. The reader should be familiar with 
description logics and the Web Ontology Language (OWL) to fully understand the construction of the 
knowledge base. An introduction to description logics from the perspective of the biomedical scientist is 
available on the wiki at {{https://github.com/loopasam/ftc/wiki/Description-Logics}}. The FTC implementation 
relies mostly on Brain \citep{Croset2013} and the web application builds on the top of the Play! framework. 
Classification tasks use the ELK reasoner \citep{Kazakov2011}. The computer hosting the web application has 8 Gb of memory 
with 4 processors, this architecture allows fast parallel reasoning, thanks to ELK’s design. More functionalities 
will be added to the web application following user requirements (lean implementation).

@@

\subsection{Categories related to biological processes}
All the biological processes featured by the GO are looked-up one by one. All the 
time a process is linked to another process (X) via a positive or negative regulation link, two FTC 
classes are created: Anti-X agent and Pro-X agent. For instance the GO term positive regulation of blood 
coagulation is linked to the term blood coagulation via a positively regulates relation, therefore two FTC 
categories Anti-blood coagulation agent and Pro-blood coagulation agent are created. The identifiers of the 
new FTC classes are also derived from the GO term used to create the class pattern. The GO numeric identifier 
is re-used and the letter A or P is appended before to emphasize the anti or pro pattern. From the example presented 
previously, the FTC class Anti-blood coagulation has FTC\_A0007596 as identifier, because the GO term blood coagulation is 
referenced by GO:0007596. Following the same logic, FTC\_P0007596 is the identifier of the class Pro-blood coagulation. The 
design choice for identifiers and labels allows the FTC to fully rely on the high quality work provided by the GO curation 
team and scale over it.

\subsection{Categories related to molecular functions}
The mode and mechanism of actions related to molecular functions are created in the following manner: All the time 
a molecular function (Y) is encountered then two FTC categories are created, as for the processes: Anti-Y agent and 
Pro-Y agent. The identifiers are assigned the same way as described before. For instance, out of the GO term androgen 
receptor activity (GO:0004882) two FTC classes are derived: Pro-androgen receptor activity agent (FTC\_P0004882) and 
Anti-androgen receptor activity agent (FTC\_A0004882).

\section{Equivalent definitions}
FTC classes are generated as presented in the previous section. Up to this point, these categories 
are only individual tokens with a human readable label as well as an identifier. The next step is going to 
assign equivalent definitions to each FTC class. A reasoner can understand such definitions and will automatically 
classify the knowledge base accordingly; drugs will then be assigned to FTC categories and the taxonomic structure 
arises after the reasoning step. Equivalent definitions are written as OWL class expressions using the entities of 
the knowledge base. The core classes and properties are presented in the supplementary material, this section assumes 
that the reader is already familiar with the FTC specifications (summarised at {{https://github.com/loopasam/ftc/wiki/Knowledge-Base}}). 
There are two types of equivalences: The first one captures perturbation of regulatory 
biological processes (so called regulatory patterns) and the second one handles the perturbed functions (functional patterns).

>>>>

\section{Source code}
The code behind the creation of the resource is entirely open and available 
at {{https://github.com/loopasam/ftc}}. The web application built on the top of the 
FTC can be find at {{https://www.ebi.ac.uk/chembl/ftc}} and the documentation can be 
accessed at {{https://github.com/loopasam/ftc/wiki}}. The reader should be familiar with 
description logics and the Web Ontology Language (OWL) to fully understand the construction of the 
knowledge base. An introduction to description logics from the perspective of the biomedical scientist is 
available on the wiki at {{https://github.com/loopasam/ftc/wiki/Description-Logics}}. The FTC implementation 
relies mostly on Brain \citep{Croset2013} and the web application builds on the top of the Play! framework. 
Classification tasks use the ELK reasoner \citep{Kazakov2011}. The computer hosting the web application has 8 Gb of memory 
with 4 processors, this architecture allows fast parallel reasoning, thanks to ELK's design. More functionalities 
will be added to the web application following user requirements (so called \emph{lean implementation}).

@@

\subsection{Categories related to biological processes}
All the biological processes featured by the GO are looked-up one by one. All the 
time a process is linked to another process (\emph{X}) via a \emph{positive} or \emph{negative regulation} link, two FTC 
classes are created: \emph{Anti-X agent} and \emph{Pro-X agent}. For instance the GO term \emph{positive regulation of blood 
coagulation} is linked to the term \emph{blood coagulation} via a \emph{positively regulates} relation, therefore two FTC 
categories \emph{Anti-blood coagulation agent} and \emph{Pro-blood coagulation agent} are created. The identifiers of the 
new FTC classes are also derived from the GO term used to create the class pattern. The GO numeric identifier 
is re-used and the letter \emph{A} or \emph{P} is appended before to emphasize the \emph{anti} or \emph{pro} pattern. 
From the example presented 
previously, the FTC class \emph{Anti-blood coagulation} has FTC\_A0007596 as identifier, because the GO term \emph{blood coagulation} is 
referenced by GO:0007596. Following the same logic, FTC\_P0007596 is the identifier of the class \emph{Pro-blood coagulation agent}. The 
design choice for identifiers and labels allows the FTC to fully rely on the high quality work provided by the GO curation 
team and scale over it.

\subsection{Categories related to molecular functions}
The mode and mechanism of actions related to molecular functions are created in the following manner: All the time 
a molecular function (\emph{Y}) is encountered then two FTC categories are created, as for the processes: \emph{Anti-Y agent} and 
\emph{Pro-Y agent}. The identifiers are assigned the same way as described before. For instance, out of the GO term \emph{androgen 
receptor activity} (GO:0004882) two FTC classes are derived: \emph{Pro-androgen receptor activity agent} (FTC\_P0004882) and 
\emph{nti-androgen receptor activity agent} (FTC\_A0004882).

\section{Equivalent definitions}
FTC classes are generated as presented in the previous section. Up to this point, these categories 
are only tokens with a human readable label as well as an identifier. The next step is going to 
assign equivalent definitions to each FTC class. A reasoner can understand such definitions and will automatically 
classify the knowledge base accordingly; drugs will then be assigned to FTC categories and the taxonomic structure 
arises after this reasoning step. Equivalent definitions are written as OWL class expressions using the entities of 
the knowledge base (summarised at {{https://github.com/loopasam/ftc/wiki/Knowledge-Base}} and in section ).

//TODO refer to section

There are two types of equivalences: The first one captures perturbation of regulatory 
biological processes (so called regulatory patterns) and the second one handles the perturbed functions (functional patterns).
